---
title: "comtrendr"
output: rmarkdown::html_vignette
lang: en
vignette: >
  %\VignetteIndexEntry{comtrendr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(comtrendr)
```

The **comtrendr** package is a suite of tools for calculating community stability metrics. It includes a framework to decompose stability into its compensatory dynamics, dominance and diversity effects, multivariate version of community stability and tools for checking data structure. It incorporates several procedures to check and account for trends in data that might lead to spurious patterns of synchrony or variance.

## Cleaning and formatting data
Most functions in `comtrendr` work with a data.frame of species data in columns and time steps in rows. The package includes the function `comm_clean()` to help with this. This function gives the appropriate format to input data in wide (time by species) and long (time and species in two separate columns) formats. In addition, it allows removing years with no data and filtering out transient species that only appear in certain number of years based on a threshold determined by the user. It is also possible to specify a column with a community identifier if we have several communities in our data or a time column. `comm_clean()` also sorts each community in chronological order if a time column is provided or assumes that data are in order if not.

```{r formatting_wide, fig.height=4, fig.width=7}
# Load example data
data("example_data_long", package = "comtrendr")

# Show first rows of the data
head(example_data_long)

# We remove species appearing in less than 30% of time steps, but keep time steps with no data.
clean_example <- comm_clean(x = example_data_long,
                input_format = "long",
                community_col = "comm", # Column with community id
                time_col = "time", # Column with time information
                empty_years = FALSE, # Remove years with no data
                filter_transient = TRUE, # Filter out 
                min_samples = 0.3) # Minimum proportion of years with data to keep a species

# Data are returned in a wide format and ready to use
head(clean_example)
```


## Data simulation

Let's work on an example using some simulated data replicating a community after a disturbance, in which all species have an overall positive trend, indicating recovery, but with asynchronous abundance changes due to different year to year environmental responses. We are going to simulate 30 years of data for a community of 10 species. This is going to be a relatively low evenness (even = 0.4), with a relation between mean abundance and variance (slope of a Taylor's power law) of 1.4 and a global positive trend (trend_mean = 0.01). We set trend_sd to 0 to ensure that the majority of species have positive trends (although some variation is introduced when the function samples from the multivariate normal distribution). We can parameterize the temporal autocorrelation between species abundances using the argument `corr`. Here we are going to set it to -0.111, the lowest value possible for a community of 10 species. Although the `corr` argument is limited between -1 and 1 because it is a correlation coefficient, in practice its lower bound depends on the number of species. The `sim_mvcomm()` function will return an error if the value set to `corr` is not valid, and it will tell you the lowest value possible for that number of species.  

```{r simulation}
set.seed(1234) # set seed for reproducibility

sim_comm <- sim_mvcomm(n_sp = 10, # number of species
                       years = 30, # length of the time series
                       even = 0.4, # evenness
                       corr = -0.111, # correlation
                       power = 1.4, # slope of Taylor's power law
                       trend_mean = 0.01, # overall slope
                       trend_sd = 0) # slope SD

# add a time column simulating that data were taken from 1991 onwards
sim_comm$sim_data <- cbind(time = seq_len(nrow(sim_comm$sim_data)) + 1990,
                           sim_comm$sim_data)

head(sim_comm$sim_data)
```

`sim_mvcomm()` returns a list with four objects: `sim_comm` is a community matrix of years by species. `baseline` is the same community matrix but with the trends removed. It can be used to check if the detrended methods work properly. `true_trend` is a named vector with the true mean trends of each species in the community, estimated from the simulated data. `params` is a named vector with the values of each parameter used in the simulation.

## Basic descriptors

Now, using the `comm_info()` function, we are going to check some basic characteristics of the community. More precisely, we are going to check the number of years, its average species richness, Shannon's index and evenness; along with identifying any dominant or transient species. Here, we consider dominant any species with an average abundance above a certain threshold, in this case those above the 70th percentile of abundance across all species (dominant_threshold = 0.7). Similarly, we consider transient species those appearing in less than a certain proportion of years. In this case, we consider a minimum of 70% of samples. `comm_info()` works for one or several communities, as long as a column in the data.frame identifying each community is properly indicated using the `community_col` argument. The `time_col` argument specifies which column indicates the order of observations. If `community_col` and `time_col` are omitted, by default `comm_info()` will return a warning indicating that it assumes that all data belongs to a single community and that rows are in chronological order.

```{r information}
comm_info(sim_comm$sim_data,
          by_timestep = FALSE,
          total = "average",
          check_dominants = TRUE,
          dominant_threshold = 0.7,
          check_transient = TRUE,
          min_samples = 0.7)
```

Again, `comm_info()` returns a list with four objects, all of them indicating the community to which that information belongs to: `diversity`, with some basic metrics of taxonomic diversity, and `dominant_taxa` and `transient_taxa` with the information about which species fall in each category. We have omitted on purpose the fourth object returned by `comm_info()`, `trends`, because we will check for any long-term trends in the community in the next step.

## Trend detection

Instead of using `comm_info()` to check for trends in species abundances we are going to use the `comm_trends()` function because it gives us greater flexibility and detail. This function includes three different methods to estimate trends at different levels of organization, two at the species level (`loglinear` and `dennis`) and one at the community level (`rda`). 

The first method, `loglinear`, estimates the average trend using linear regression with log-transformed abundances as the response variable and time as the explanatory variable. Here, time is transformed so that the first time step corresponds to 0 and subsequent time steps are 1, 2, etc. This way, the estimated slope corresponds to the average change per time step.  The second method, `dennis`, calculates the log-ratios between consecutive time steps using the method of Dennis et al. (1991), which takes into account any gaps between time steps. The third method, `rda`, estimates the community-level trends using redundancy analysis on the community matrix using time as an explanatory variable.

If methods `loglinear` or `dennis` are used, `comm_info()` returns a data.frame with the average trend of each species in the community (or communities), its 95% confidence interval and *p-value*. Method `rda` returns the *F-statistic* and *p-value* of the redundandy analysis for each community. 
`comm_info()` includes a plotting option to visualize trends.

Let's check the species-level trends first using the `loglinear` method: 
```{r trends, fig.height=4, fig.width=7}
comm_trends(sim_comm$sim_data, 
            method = "loglinear", # estimate trend using linear regression on log abundances
            plot = T) # plot the results
```

We can see clearly that species 1 to 5 and 10 have significant trends, species 6 to 7 are almost significant and species 8 and 9 do not show a clear pattern. This deviation from the imputed parameters happens because each years abundances are randomly drawn from a multivariate normal distribution with means determined by the underlying trends and variance determined by the slope of a Taylor's power law (argument `power` in `sim_mvcomm()`).

If we check the community level trends using redundancy analysis by setting `method` to `rda` we can confirm that there is a statistically significant trend in the community as a whole. In the left panel, each grey dot corresponds to one time step, and the x-axis represents time. If consecutive dots follow the axis from left to right, it indicates an overall positive trend, whereas the opposite pattern indicates an overall decrease in abundance. The colored dots correspond to each species, and their arrows indicate an increase (right) of decrease (left) of abundance. The *F-statistic* and *p-value* are also shown above the left panel. 

```{r trends_multi, fig.height=4, fig.width=7}
comm_trends(sim_comm$sim_data, 
            method = "rda", # estimate trend using redundancy analysis
            plot = T) # plot the results
```

## Estimating community stability metrics

Since there are clear trends in abundance, it seems reasonable to use detrended methods for estimating community stability. For comparison's sake we are going to calculate both standard and detrended metrics. We can estimate stability and synchrony indices one by one using their respective functions `cv_com_term()` ($CV_{com}$ and $CV_{pop}$), `eta_gross()` ($\eta$, Gross et al. 2014 and Blüthgen et al. 2016), `psi_segrestin()` ($\psi$, Segrestin et al. 2024), `phi_loreau()` ($\phi$, Loreau \& de Mazancourt 2008) and `logvar_ratio()` (Lepš et al. 2018). However, it is usually more convenient to calculate all of them with the `comm_stability()` function. This returns a data.frame with these metrics for each community, whether they are detrended or not and, for $CV_{pop}$ and $\eta$, if the index value was weighted by the relative average abundance of each species.

It is important to note that standard and detrended versions of these metrics are not mutually exclusive and give different information. Standard metrics represent the overall synchrony of the community while detrended version focus on year to year variability. In communities with small or nonexistent trends, both metrics yield very similar results.

```{r stability, fig.height=4, fig.width=5}
# Compute some classic synchrony metrics
classic_sync <- comm_stability(sim_comm$sim_data, term = "var")

# Compute detrended synchrony metrics
det_sync <- comm_stability(sim_comm$sim_data, term = "two")

plot(x = seq_along(classic_sync$value), y = classic_sync$value, 
     col = 1, 
     pch = 1,
     xaxt = "n",
     ylab = "Value",
     xlab = "Stability or synchrony metric",
     ylim = range(c(classic_sync$value, det_sync$value)))
abline(h = 0, lty = 2)
points(x = seq_along(det_sync$value), y = det_sync$value,
       col = 1,
       pch = 19)
axis(1, at = 1:6, labels = c(expression(psi), expression(phi), expression(eta), 
                             "logvar", expression('CV'[com]), expression('CV'[pop])))
legend("bottomleft", legend = c("Classic", "Detrended"), col = 1, pch = c(1, 19))
```

Here, $\psi$, $\phi$, $\eta$ and $logvar$ are synchrony indices, and $CV_{com}$ and $CV_{pop}$ correspond, respectively, to the coefficeint of variation of the sum of annual abundances and the average CV across species. $\psi$ and $\phi$ are bounded between 0 and 1, with 1 indicating perfect synchrony and 0 perfect asynchrony. $\eta$ is bounded between -1 and 1, with positive values indicate synchrony, negative ones asynchrony and 0 indicating independent fluctuations. The same applies for $logvar$, although its upper limit is the natural logarithm of the number of species in the community $ln(S)$ and its lower limit equals $-\infty$.

Both CVs are reduced when using detrended variance estimates, an expected result since trends increase the variance of the time series. For stability, standard indices were located above the midpoint of their ranges, indicating a small degree of synchronous fluctuations between species. This is expected given that all species are increasing their abundances at the same time. However, using the detrended estimates reveals the underlying, year to year asynchrony that we simulated and that was masked by the overarching trends.

## Partitioning of community stability

The processes that stabilize a community (*i.e.* the difference between the CV of the community as a whole vs the CVs of individual species) can be partitioned into three main effects: dominance, asynchrony and averaging effect. The `comm_partitioning()` implements this framework, with the ability of using standard and detrended methods. Let's apply it our simulated community and plot the results:

```{r partitioning, fig.height=5, fig.width=10}
comstab_var <- comm_partitioning(sim_comm$sim_data)
comstab_two <- comm_partitioning(sim_comm$sim_data, term = "two")
comstab <- list(comstab_var, comstab_two)
class(comstab) <-  c("comstab", "comstab_list") # add class to object so that plot method works properly
plot(comstab)

```

The left panel shows the absolute effects of those stabiliziing processes. $CV_{e}$ is the coefficient of variation of an average species, and represents the expected stability of the community under no stabilization. $CV_{com}$ is the CV of the sum of annual abundances, which incorporates all the stabilizing effect. $CVtilde$ is equivalent to the weighted $CV_{pop}$ that we calculated before. Since the weighting gives more importance to the most dominant species, the difference between $CV_{e}$ and $CVtilde$ represents the stabilizing effect of those species. $CV_{a}$ corresponds to the stability of the community after removing the effect of asynchronous dynamics between species. The difference between $CV_{a}$ and $CV_{com}$ corresponds to the stabilizing effect caused by the remaining component, the averaging effect. This represents the statistical phenomenon in which more diverse communitites tend to be more stable because there is a higher chance of species compensating one another (Doak et al. 1998). Together with asynchrony, their combined effect represents the compensatory dynamics of the community. The right panel shows the relative importance of the dominance, asynchrony and averaging effects.

At first glance we see that all CVs estimated using detrended methods (red dots and lines) are lower than the standard methods (in black). This is consistent with our previous results and the fact that trends lead to overstimation of variance if they are not accounted for. In absolute terms, the dominance effect is similar with standard and detrended methods. However, the overall stabilization effect is higher using detrended variance estimates and thus the relative effect of dominance is around a 10% lower. Thus, compensatory dynamics gain importance after removing those trends, with asynchrony being the most prominent stabilizing mechanism. Again, this is consistent with our simulations and previous results, in which the overall positive trend in the community masks the short-term asynchrony between species.

## Multivariate stability

Trends lead to changes in abundance and thus in the composition of the community as a whole. We saw this when checking for community-level trends using the `rda` method. As such, it is interesting to check if not only the abundance of the community is stable overtime but also its composition To this end, `comtrendr` incorporates a multivariate version of CV available in the `cv_mv()` and `comm_stability()` (setting the argument `composition = TRUE`) functions. The multivariate variance of community composition is defined as the average squared euclidean or chord distance between annual observations and the average community composition. Its detrended version follows a similar approach but distances are calculated between consecutive years instead.

```{r composition, fig.height=5, fig.width=7}
cv_mv(sim_comm$sim_data)
cv_mv(sim_comm$sim_data, term = "two")
```

## References

- Blüthgen, N., Simons, N. K., Jung, K., Prati, D., Renner, S. C., Boch, S., ... & Gossner, M. M. (2016). Land use imperils plant and animal community stability through changes in asynchrony rather than diversity. Nature Communications, 7(1), 10697.
- Dennis, B., Munholland, P. L., & Scott, J. M. (1991). Estimation of Growth and Extinction Parameters for Endangered Species. Ecological Monographs, 61(2), 115–143.
- Doak, D. F., Bigger, D., Harding, E. K., Marvier, M. A., O’Malley, R. E., & Thomson, D. (1998). The Statistical Inevitability of Stability‐Diversity Relationships in Community Ecology. The American Naturalist, 151(3), 264-276.
- Gross, K., Cardinale, B. J., Fox, J. W., Gonzalez, A., Loreau, M., Wayne Polley, H., ... & van Ruijven, J. (2014). Species richness and the temporal stability of biomass production: a new analysis of recent biodiversity experiments. The American Naturalist, 183(1), 1-12.
- Lepš, J., Májeková, M., Vítová, A., Doležal, J., & de Bello, F. (2018). Stabilizing effects in temporal fluctuations: Management, traits, and species richness in high‐diversity communities. Ecology, 99(2), 360-371.
- Lepš, J., Götzenberger, L., Valencia, E. & de Bello, F. (2019). Accounting for long-term directional trends on year to year synchrony in species fluctuations. Ecography, 42, 1728-1741.
- Loreau, M., & de Mazancourt, C. (2008). Species synchrony and its drivers: neutral and nonneutral community dynamics in fluctuating environments. The American Naturalist, 172(2), E48-E66.
- Segrestin, J., Götzenberger, L., Valencia, E., de Bello, F., & Lepš, J. (2024). A unified framework for partitioning the drivers of stability of ecological communities. Global Ecology and Biogeography, 33(5), e13828.

