---
title: "worked_example"
output: rmarkdown::html_vignette
author: Héctor Miranda-Cebrián
vignette: >
  %\VignetteIndexEntry{worked_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(comtrendr)
```

## Data simulation
Let's work on an example using some simulated data replicating a community after a disturbance, in which all species have an overall positive trend, indicating recovery, but with asynchronous abundance changes due to different year-to-year environmental responses. We are going to simulate 30 years of data for a community of 10 species. This is going to be a relatively low evenness (even = 0.4), with a relation between mean abundance and variance (slope of a Taylor's power law) of 1.4 and a global positive trend (trend_mean = -0.01). We set trend_sd to 0 to ensure that the majority of species have positive trends (although some variation is introduced when the function samples from the multivariate normal distribution). We can parameterize the temporal autocorrelation between species abundances using the argument `corr`. Here we are going to set it to -0.111, the lowest value possible for a community of 10 species. Although the `corr`argument is limited between -1 and 1 because it is a correlation coefficient, in practice its lower bound depends on the number of species. The `sim_mvcomm()`function will return an error if the value set to `corr` is not valid, and it will tell you the lowest value possible for that number of species.  

```{r simulation}
library(comtrendr)

set.seed(1234) # set seed for reproducibility

sim_comm <- sim_mvcomm(n_sp = 10, # number of species
                       years = 30, # length of the time series
                       even = 0.4, # evenness
                       corr = -0.111, # correlation
                       power = 1.4, # slope of Taylor's power law
                       trend_mean = 0.01, # overall slope
                       trend_sd = 0) # slope SD

# add a time column simulating that data were taken from 1991 onwards
sim_comm$sim_data <- cbind(time = seq_len(nrow(sim_comm$sim_data)) + 1990,
                           sim_comm$sim_data)

head(sim_comm$sim_data)
```

`sim_mvcomm()` returns a list with four objects: `sim_comm` is a community matrix of years by species. `baseline` is the same community matrix but with the trends removed. It can be used to check if the detrended methods work properly. `true_trend` is a named vector with the true mean trends of each species in the community, estimated from the simulated data. `params` is a named vector with the values of each parameter used in the simulation.

## Basic descriptors

Now, using the `comm_info()` function, we are going to check some basic characteristics of the community. More precisely, we are going to check the number of years, its average species richness, Shannon's index and evenness; along with identifying any dominant or transient species. Here, we consider dominant any species with an average abundance above a certain threshold, in this case those above the 70th percentile (dominant_threshold = 0.7). Similarly, we consider transient species those appearing in less than a certain proportion of years. In this case, we consider a minimum of 70% of samples. `comm_info()` works for one or several communities, as long as a column in the data.frame identifying each community is properly indicated using the `community_col` argument. The `time_col`argument specifies which column indicates the order of observations. If `community_col` and `time_col` are ommited, by default `comm_info()` will return a warning indicating that it assumes that all data belongs to a single community and that rows are in chronological order.

```{r information}
comm_info(sim_comm$sim_data,
          by_timestep = FALSE,
          total = "average",
          check_dominants = TRUE,
          dominant_threshold = 0.7,
          check_transient = TRUE,
          min_samples = 0.7)
```

Again, `comm_info()` returns a list with four objects, all of them indicating the community to which that information belongs to: `diversity`, with some basic metrics of taxonomic diversity, and `dominant_taxa` and `transient_taxa` with the information about which species fall in each category. We have ommited on purpose the fourth object returned by `comm_info()`, `trends`, because we will check for any long-term trends in the community in the next step.

## Trend detection

Instead of using `comm_info()` to check for trends in species abundances we are going to use the `comm_trends()` function because it gives us greater flexibility and detail. This function includes three different methods to estimate trends at different levels of organization, two at the species level (`loglinear` and `dennis`) and one at the community level (`rda`). 

The first method, `loglinear`, estimates the average trend using linear regression with log-transformed abundances as the response variable and time as the explanatory variable. Here, time is transformed so that the first timestep corresponds to 0 and subsequent timesteps are 1, 2, etc. This way, the estimated slope corresponds to the average change per timestep.  The second method, `dennis`, calculates the log-ratios between consecutive timesteps using the method of Dennis et al. (1993), which takes into account any gaps between timesteps. The third method, `rda`, estimates the community-level trends using redundancy analysis on the community matrix using time as an explanatory variable.

If methods `loglinear`or `dennis` are used, `comm_info()`returns a data.frame with the average trend of each species in the community (or communities), its 95% confidence interval and *p-value*. Method `rda` returns the *F-statistic* and *p-value* of the redundandy analysis for each community. 
`comm_info()` includes a plotting option to visualize trends.

Let's check the species-level trends first using the `loglinear` method: 
```{r trends, fig.height=4, fig.width=7}
comm_trends(sim_comm$sim_data, 
            method = "loglinear", # estimate trend using linear regression on log abundances
            plot = T) # plot the results
```

We can see clearly that species 1 to 5 and 10 have significant trends, species 6 to 7 are almost significant and species 8 and 9 do not show a clear pattern. This deviation from the inputed parameters for the simulation is caused by the samping process from the multivariate normal distribution, because each years abundances are randomly drawn from a distribution with means determined by the underlying trends and variance determined by the slope of a Taylor's power law (argument `power` in `sim_mvcomm()`).

If we check the community level trends using redundancy analysis by setting `method` to `rda` we can confirm that there is a statistically significant trend in the community as a whole. In the left panel, each grey dot corresponds to one timestep, and the x-axis represents time. If consecutive dots follow the axis from left to right, it indicates an overall positive trend, whereas the opposite pattern indicates an overall decrease in abundance. The colored dots correspond to each species, and their arrows indicate the direction of abundance change for that particular species. The *F-statistic* and *p-value* are also shown above the right panel. 

```{r trends_multi, fig.height=4, fig.width=7}
comm_trends(sim_comm$sim_data, 
            method = "rda", # estimate trend using redundancy analysis
            plot = T) # plot the results
```

Since there is a clear trend in abundance, it seems reasonable to use detrended methods for estimating community stability. But first, we are going to remove species 7 from our data using the `comm_clean()` function because it is missing many years and thus could bias our results.

```{r clean_data}
clean_data <- comm_clean(sim_comm$sim_data,
                         input_format = "wide",
                         filter_transient = TRUE,
                         min_samples = 0.3)
```

## Estimating community stability metrics

Since there are . We can estimate stability and synchrony indices one by one using their functions `cv_com_term()`, `eta_gross()`, `psi_segrestin()`, `phi_loreau()` and `logvar_ratio()`. However, it is usually more convenient to calculate all of them with the `comm_stability()`function. This returns a data.frame with these metrics for each community, whether they are detrended or not and, for \eqn{CV_{pop}} and \eta if the index value was weighted by the average relative abundance of each species.

```{r stability, fig.height=4, fig.width=5}
# Compute some classic synchrony metrics
classic_sync <- comm_stability(clean_data, term = "var")

# Compute detrended synchrony metrics
det_sync <- comm_stability(clean_data, term = "two")

plot(x = seq_along(classic_sync$value), y = classic_sync$value, 
     col = 1, 
     pch = 1,
     xaxt = "n",
     ylab = "Value",
     xlab = "Stability or synchrony metric",
     ylim = c(min(c(classic_sync$value, det_sync$value)), 1))
abline(h = 0, lty = 2)
abline(h = 1, lty = 1)
points(x = seq_along(det_sync$value), y = det_sync$value,
       col = 1,
       pch = 19)
axis(1, at = 1:6, labels = c(expression(psi), expression(phi), expression(eta), 
                             "logvar", expression('CV'[com]), expression('CV'[pop])))
legend("bottomleft", legend = c("Classic", "Detrended"), col = 1, pch = c(1, 19))
```

## Partitioning of community stability

## Multivariate stability